Sparse spectral and $p$-finite element methods for partial differential equations on disk-slices and trapeziums

### 2019-04-17 papers/halfdisk/halfdisk.tex
New figs. Some comments. New ref.

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs on a half-sphere can be represented using two families of OPs on the half-disk \sotodo{Add citation to this} \bstodo{should we instead take the Xu approach?}, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family. Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in fluids.

@@

\begin{figure}
	\begin{subfigure}[t]{0.3\textwidth}
	\centering
	\includegraphics[scale=0.3]{solution-poisson}
	\caption{}
	\label{fig:solution-poisson}
	\end{subfigure}
	\begin{subfigure}[t]{0.3\textwidth}
	\centering
	\includegraphics[scale=0.3]{solution-helmholtz}
	\caption{}
	\label{fig:solution-helmholtz}
	\end{subfigure}
	\begin{subfigure}[t]{0.3\textwidth}
	\centering
	\includegraphics[scale=0.3]{solution-biharmonic}
	\caption{}
	\label{fig:solution-biharmonic}
	\end{subfigure}
	\centering
    	\caption{(a): The computed solution to $\Delta u = f$ with zero boundary conditions with $f(x,y) = 1 + \text{erf}(5(1 - 10((x - 0.5)^2 + y^2)))$. (b): The computed solution to $\Delta u + k^2 \: v \: u = f$ with zero boundary conditions with $f(x,y) = xye^x$, $v(x,y) = xy^2$ and $k = 200$. (c): The computed solution to $\Delta^2 u = f$ with zero Dirichlet and Neumann boundary conditions with $f(x,y) = 1 + \text{erf}(5(1 - 10((x - 0.5)^2 + y^2)))$.}
        \label{fig:solution}
\end{figure}

@@

We can tackle the problem as follows. Fix \(a, b = 1,1\). Denote the coefficient vector for expansion of $u$ in the $\bigWii$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bighdopii$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain $\mathbf{f}$ using the quadrature rule above. In matrix-vector notation, our system hence becomes:
\begin{align}
    \laplacewii \mathbf{u} = \mathbf{f}
\end{align}
which can be solved to find $\mathbf{u}$.
In Figure \ref{fig:solution-poisson} we see the solution to the Poisson equation with zero boundary conditions given in (\ref{eqn:poisson}) in the half-disk $\Omega$. In Figure \ref{fig:solutionblocknorms} we show the norms of each block of calculated coefficients of the approximation for four right-hand sides of the Poisson equation with N = 200, that is, 20,301 unknowns. The rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution. We see that... \bstodoinline{Describe what plot tells us - I dont want to copy and paste your Triangle paper here but not sure what else to write}

\begin{figure}
        	\includegraphics[scale=0.5]{solutionblocknorms}
	\centering
	\caption{The norms of each block of the computed solution of the Poisson equation with the given right hand side functions. This demonstrates algebraic convergence with the rate dictated by the decay at the corners, with spectral convergence observed when the right-hand side vanishes to all orders.  \bstodoinline{N=200 blocks here. When I have the time to run N=1000 or so I will do}}
        	\label{fig:solutionblocknorms}
\end{figure}

@@

\sotodoinline{Rewrite in a more "paper" like tone}

>>>>

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs on a half-sphere can be represented using two families of OPs on the half-disk \cite[Theorem 3.1]{olver2018orthogonal}, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family. Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in fluids.

@@

\begin{figure}
	\begin{subfigure}[t]{0.3\textwidth}
	\centering
	\includegraphics[scale=0.3]{solution-poisson}
	\caption{}
	\label{fig:solution-poisson}
	\end{subfigure}
	\begin{subfigure}[t]{0.3\textwidth}
	\centering
	\includegraphics[scale=0.3]{solution-helmholtz}
	\caption{}
	\label{fig:solution-helmholtz}
	\end{subfigure}
	\begin{subfigure}[t]{0.3\textwidth}
	\centering
	\includegraphics[scale=0.3]{solution-biharmonic}
	\caption{}
	\label{fig:solution-biharmonic}
	\end{subfigure}
	\centering
    	\caption{(a): The computed solution to $\Delta u = f$ with zero boundary conditions with $f(x,y) = 1 + \text{erf}(5(1 - 10((x - 0.5)^2 + y^2)))$. (b): The computed solution to $\Delta u + k^2 \: v \: u = f$ with zero boundary conditions with $f(x,y) = xye^x$, $v(x,y) = 1 - (3(x-1)^2 + 5y^2)$ and $k = 200$. (c): The computed solution to $\Delta^2 u = f$ with zero Dirichlet and Neumann boundary conditions with $f(x,y) = 1 + \text{erf}(5(1 - 10((x - 0.5)^2 + y^2)))$.}
        \label{fig:solution}
\end{figure}

@@

We can tackle the problem as follows. Fix \(a, b = 1,1\). Denote the coefficient vector for expansion of $u$ in the $\bigWii$ OP basis up to degree $N$ by $\mathbf{u}$, and the coefficient vector for expansion of $f$ in the $\bighdopii$ OP basis up to degree $N$ by $\mathbf{f}$. Since $f$ is known, we can obtain $\mathbf{f}$ using the quadrature rule above. In matrix-vector notation, our system hence becomes:
\begin{align}
    \laplacewii \mathbf{u} = \mathbf{f}
\end{align}
which can be solved to find $\mathbf{u}$.
In Figure \ref{fig:solution-poisson} we see the solution to the Poisson equation with zero boundary conditions given in (\ref{eqn:poisson}) in the half-disk $\Omega$. In Figure \ref{fig:solutionblocknorms} we show the norms of each block of calculated coefficients of the approximation for four right-hand sides of the Poisson equation with N = 200, that is, 20,301 unknowns. The rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution. We see that we achieve algebraic convergence for the first three examples, with the final Gaussian bump example achieving spectral convergence. We note that for right hand sides that vanish at the corners of our half disk ($y = \pm 1$) we see faster convergence. \bstodoinline{Improve above}

\begin{figure}
        	\includegraphics[scale=0.5]{solutionblocknorms}
	\centering
	\caption{The norms of each block of the computed solution of the Poisson equation with the given right hand side functions. This demonstrates algebraic convergence with the rate dictated by the decay at the corners, with spectral convergence observed when the right-hand side vanishes to all orders.  \bstodoinline{N=200 blocks here}}
        	\label{fig:solutionblocknorms}
\end{figure}

@@

\bstodoinline{Rewritten}
